---
title: "P Metabolism"
author: "JAW"
date: '`r format(Sys.time(), "%H:%M:%S on %d %B, %Y")`'
output: 
  html_document:
    number_sections: TRUE
    toc: TRUE
    toc_float: TRUE
---

# Set up

```{r}
library("asreml")
```

```{r}
qgm_summary <- function(asr,wald) {
  if(missing(wald)) {wald <- TRUE}
  print("CONVERGENCE")
  print(paste0(asr$converge," with logL = ",asr$loglik))
  cat("\n")
  print("VARIANCE COMPONENTS")
  print(summary(asr)$varcomp)
  cat("\n")
  print("FIXED EFFECTS")
  print(as.data.frame(summary(asr,coef=TRUE)$coef.fixed))
  if(wald) {
    cat("\n")
    print("WALD TESTS")
    print(as.data.frame(wald(asr,denDF="numeric",ssType="conditional")$Wald)[,c(1,2,4,6)])
  }
}
```

```{r}
qgm_continue <- function(asr, iter, foll) {
  if(missing(foll)) {foll = FALSE}
  asreml.options(extra = iter, trace = foll)
  asr <- update.asreml(asr)
  asreml.options(extra = 0, trace = TRUE)
  return(asr)
}
```

# Data

-   `y1` has no block variance
-   `y2` is modified from `y1` with large block variance by
    -   `y2`=`y1`-8 for block 1,
    -   `y2`=`y1`+8 for block 2,
    -   `y2`=`y1` for block 3

```{r}
hay <- data.frame(treatment= c(rep(rep(c(1,2),each=4),times=2),rep(3,times=8)), block= rep(c(1,2,3),each=8), y1= c(28,26,25,25,23,24,24,21,28,25,27,26,25,23,24,24,38,37,39,37,39,39,39,38), y2= c(20,18,17,17,15,16,16,13,36,33,35,34,33,31,32,32,38,37,39,37,39,39,39,38))
hay
nrow(hay)
```

# Design

-   Confounding of treatment 3 with block 3
-   Blocks 1 & 2, and treatments 1 & 2 balanced

```{r}
addmargins(table(hay$treatment,hay$block))
```

# Fixed `block` ... fails

-   `asreml()` produces an answer but after identifying confounding
    -   here `block 3` is set to precisely 0
        -   s.e. is `NA`
    -   `treatment 3` get a value, but only because the `block 3` set to 0
        -   this is because `treatment` appears first in the model line
        -   the only justification for this is to provide an answer, and leave the user to interpret the mess!
-   behaviour is the same as `lm()`!

## Using `asreml()`

```{r}
hay$treatment <- factor(hay$treatment)
hay$block <- factor(hay$block)
asr <- asreml(fixed= y1 ~1 +treatment +block, residual= ~units, data=hay)
qgm_summary(asr)
```

## Using `lm()`

-   same answers

```{r}
summary(lm(y1 ~1 +treatment +block, dat=hay))
```

# Random `block` with `y1`

-   Small block variance
    -   LRT is (P\>0.05)!
-   `treatment 3` identified as significantly increasing `y1`
    -   ... as observing such a large block effect is considered very unlikely

```{r}
asr <- asreml(fixed= y1 ~1 + treatment, residual= ~units, data=hay)
asr <- asreml(fixed= y1 ~1 + treatment, random= ~block, residual= ~units, data=hay)
asr <- qgm_continue(asr,5)
qgm_summary(asr)
```

# Random `block` with `y2`

-   Large block variance
    -   LRT indicates clearly \>0
    -   note how the notional 1 d.f. for estimating the block variance (between blocks 1 & 2) gives considerable uncertainty as to how large the variance actually is other than it is \>0!
-   `y2` has same fixed effects and same residual variance
    -   the data were constructed to be just that for demonstration purposes, and normally this is not so!
-   ... now it is considered far less clear that`treatment 3` has an effect
    -   as a large block effect cannot be considered unlikely

```{r}
asr <- asreml(fixed= y2 ~1 + treatment, residual= ~units, data=hay)
asr <- asreml(fixed= y2 ~1 + treatment, random= ~block, residual= ~units, data=hay)
asr <- qgm_continue(asr,5)
qgm_summary(asr)
```
