---
title: "Attempted Reproduction of Scotland CMS PDF Analyses"
author: "Junxiu"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
params:
  run_heavy_steps: false
  n_replicates: 100
  n_candidates: 700
---

# Scope

This notebook is an **attempted reproduction** of the PDF:

-   `ScotlandCMS_ISGA_PSLv4_MB.pdf`

It follows the same structure:

1.  Validation with qPCR / histopathology
2.  qPCR GWAS
3.  Simulation of selection response (qPCR index, histopathology index, combined index)

Because the PDF does not include all implementation details (for example, exact model formulas and all QC thresholds), some parts are reconstructed with explicit assumptions.

# Setup

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5
)

set.seed(20260312)
run_heavy_steps <- isTRUE(params$run_heavy_steps)
```

```{r packages}
required_pkgs <- c(
  "tidyverse",
  "readxl",
  "janitor",
  "stringr",
  "genio",
  "MASS",
  "Matrix"
)

missing_pkgs <- required_pkgs[
  !vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_pkgs) > 0) {
  stop(
    "Missing packages: ", paste(missing_pkgs, collapse = ", "),
    "\nPlease install them before knitting."
  )
}

invisible(lapply(required_pkgs, library, character.only = TRUE))
```

# File Paths

```{r paths}
# Phenotype
pheno_file1 <- file.path(
  "data", "phenotype",
  "CMS challenge Scotland 22G BE20041B sample IDs_with PCR11102022.xlsx"
)
pheno_file2 <- file.path(
  "data", "phenotype",
  "Data - CMS MOWI genetics project Scotland.xlsx"
)

# Pedigree
ped_file <- file.path("data", "pedigree", "pedigree22G_CMSSalmoStrong.txt")

# Genotype
bed_file <- file.path("data", "genotype", "2022CMSGeno_SalmoStrong.bed")
bim_file <- file.path("data", "genotype", "2022CMSGeno_SalmoStrong.bim")
fam_file <- file.path("data", "genotype", "2022CMSGeno_SalmoStrong.fam")
plink_prefix <- tools::file_path_sans_ext(bed_file)
```

# Helper Functions

```{r helpers}
numify <- function(x) suppressWarnings(as.numeric(x))

normalize_cms_id <- function(x) {
  x <- as.character(x)
  x[x %in% c("", "NA", "NaN", "0", "2")] <- NA_character_

  # Some IDs in the workbook have missing zero blocks.
  x <- sub("^2022IRS1ICMFR", "2022IRS1ICM00FR", x)
  x <- sub("^2018IRS1CBN([0-9]{5}[A-Z][0-9]{2})$", "2018IRS1CBN0000\\1", x)
  x
}

build_histo_key_from_cassette <- function(x) {
  x <- as.character(x)
  has_prefix <- !is.na(x) & stringr::str_detect(x, "^\\d{4}/\\d{4}")

  out <- rep(NA_character_, length(x))
  out[has_prefix] <- paste0(
    stringr::str_replace_all(stringr::str_extract(x[has_prefix], "^\\d{4}/\\d{4}"), "/", "_"),
    "_",
    stringr::str_pad(as.integer(stringr::str_extract(x[has_prefix], "\\d+$")), width = 3, pad = "0")
  )
  out
}

build_histo_key_from_histo_id <- function(x) {
  x <- as.character(x)
  paste0(
    stringr::str_extract(x, "\\d{4}_\\d{4}"),
    "_",
    stringr::str_pad(as.integer(stringr::str_extract(x, "(?<=_)\\d+(?=\\s*\\()")), width = 3, pad = "0")
  )
}

safe_cv <- function(x) {
  100 * sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE)
}
```

# Data Import

```{r import}
sheet1 <- readxl::read_excel(pheno_file1, sheet = "Sheet1") %>%
  janitor::clean_names()

sheet2 <- readxl::read_excel(pheno_file1, sheet = "Sheet2") %>%
  janitor::clean_names()

histo_raw <- readxl::read_excel(pheno_file2, sheet = "Scotland samples", skip = 1) %>%
  janitor::clean_names()

pedigree <- readr::read_table2(ped_file) %>%
  mutate(
    sire = na_if(as.character(sire), "0"),
    dam = na_if(as.character(dam), "0")
  )

fam <- readr::read_table2(
  fam_file,
  col_names = c("fid", "iid", "pid", "mid", "sex", "plink_pheno")
)

bim <- readr::read_table2(
  bim_file,
  col_names = c("chr", "snp", "cm", "bp", "a1", "a2")
)
```

# Rebuild Validation Dataset

```{r phenotypes}
# qPCR from Sheet1 (used for ID linkage to genotype).
pheno_qpcr <- sheet1 %>%
  transmute(
    id = normalize_cms_id(animal),
    cp_1 = numify(cp_value_12),
    cp_2 = numify(cp_value_13),
    cp_3 = numify(cp_value_14)
  ) %>%
  distinct(id, .keep_all = TRUE) %>%
  mutate(
    cp_1 = na_if(cp_1, 0),
    cp_2 = na_if(cp_2, 0),
    cp_3 = na_if(cp_3, 0),
    ct = rowMeans(cbind(cp_1, cp_2, cp_3), na.rm = TRUE),
    ct = if_else(is.nan(ct), NA_real_, ct)
  )

# qPCR from Sheet2 (used to align with sample_id records from challenge batch).
pcr_sheet2 <- sheet2 %>%
  transmute(
    sample_id = as.character(sample_id),
    cp_1 = numify(cp_value_8),
    cp_2 = numify(cp_value_9),
    cp_3 = numify(cp_value_10),
    ct = rowMeans(cbind(cp_1, cp_2, cp_3), na.rm = TRUE),
    ct = if_else(is.nan(ct), NA_real_, ct)
  )

# Mapping table from sample_id to breeding IDs and histopathology key.
sample_map <- sheet1 %>%
  transmute(
    sample_id = as.character(sample_id),
    id = normalize_cms_id(animal),
    sire = normalize_cms_id(sire),
    dam = normalize_cms_id(dam),
    histo_key = build_histo_key_from_cassette(histo_cassette_id)
  ) %>%
  distinct(sample_id, .keep_all = TRUE)

# Histopathology table.
pheno_histo <- histo_raw %>%
  transmute(
    histo_id = as.character(id),
    histo_key = build_histo_key_from_histo_id(id),
    atrium_pct_inflam = numify(percentage_inflammation),
    spong_pct_inflam = numify(precentage_inflammation),
    ratio_inflam = numify(ratio_inflammation_atrium_spongiosum)
  )

# Main validation dataset (qPCR + histopathology overlap through sample map).
validation_dat <- pcr_sheet2 %>%
  left_join(sample_map, by = "sample_id") %>%
  inner_join(pheno_histo, by = "histo_key")
```

```{r summary_stats}
qpcr_summary <- pheno_qpcr %>%
  summarise(
    n = sum(!is.na(ct)),
    mean = mean(ct, na.rm = TRUE),
    sd = sd(ct, na.rm = TRUE),
    min = min(ct, na.rm = TRUE),
    max = max(ct, na.rm = TRUE),
    cv_percent = safe_cv(ct)
  )

histo_summary <- pheno_histo %>%
  summarise(
    n = n(),
    atrium_mean = mean(atrium_pct_inflam, na.rm = TRUE),
    atrium_sd = sd(atrium_pct_inflam, na.rm = TRUE),
    spong_mean = mean(spong_pct_inflam, na.rm = TRUE),
    spong_sd = sd(spong_pct_inflam, na.rm = TRUE)
  )

overlap_summary <- validation_dat %>%
  summarise(
    n_overlap = n(),
    cor_ct_ratio = cor(ct, ratio_inflam, use = "pairwise.complete.obs"),
    cor_ct_atrium = cor(ct, atrium_pct_inflam, use = "pairwise.complete.obs"),
    cor_ct_spong = cor(ct, spong_pct_inflam, use = "pairwise.complete.obs")
  )

qpcr_summary
histo_summary
overlap_summary
```

```{r basic_plots}
pheno_qpcr %>%
  ggplot(aes(x = ct)) +
  geom_histogram(bins = 35, fill = "#2C7FB8", color = "white") +
  labs(
    title = "Distribution of qPCR CT (Sheet1)",
    x = "CT value",
    y = "Count"
  ) +
  theme_bw()

validation_dat %>%
  ggplot(aes(x = ct, y = ratio_inflam)) +
  geom_point(alpha = 0.6, color = "#1D3557") +
  geom_smooth(method = "lm", se = TRUE, color = "#E63946") +
  labs(
    title = "qPCR CT vs Histopathology Inflammation Ratio",
    x = "qPCR CT",
    y = "Inflammation ratio (atrium/spongiosum)"
  ) +
  theme_bw()
```

# Pedigree and Genotype Coverage

```{r coverage}
family_counts <- sample_map %>%
  filter(!is.na(sire), !is.na(dam)) %>%
  count(sire, dam, name = "n_offspring")

coverage_table <- tibble(
  metric = c(
    "Pedigree individuals",
    "Genotyped individuals (.fam)",
    "SNP markers (.bim)",
    "qPCR records with normalized breeding ID",
    "Genotyped + qPCR overlap",
    "Observed sire-dam families in mapped samples"
  ),
  value = c(
    nrow(pedigree),
    nrow(fam),
    nrow(bim),
    sum(!is.na(pheno_qpcr$id) & !is.na(pheno_qpcr$ct)),
    sum(fam$iid %in% pheno_qpcr$id[!is.na(pheno_qpcr$ct)]),
    nrow(family_counts)
  )
)

coverage_table
```

# Optional: Heritability Models (ASReml)

```{r asreml_example, eval=FALSE}
# This chunk is optional because ASReml requires a separate licensed installation.
# The model structure mirrors the pedigree/genomic heritability idea in the PDF.

library(asreml)

sire_dat <- sample_map %>%
  select(id, sire) %>%
  left_join(pheno_qpcr %>% select(id, ct), by = "id") %>%
  filter(!is.na(ct), !is.na(sire))

fit_sire <- asreml(
  fixed = ct ~ 1,
  random = ~ sire,
  residual = ~ idv(units),
  data = sire_dat
)

summary(fit_sire)$varcomp
```

# Optional Heavy Step: GWAS on qPCR CT

```{r gwas}
if (!run_heavy_steps) {
  cat("GWAS is skipped. Set params$run_heavy_steps = TRUE to run it.\n")
} else {
  matched <- fam %>%
    mutate(row_index = row_number()) %>%
    transmute(id = iid, row_index) %>%
    left_join(pheno_qpcr %>% select(id, ct), by = "id") %>%
    filter(!is.na(ct))

  # Read genotype matrix (SNP x individual).
  plink_obj <- genio::read_plink(plink_prefix)
  G <- plink_obj$X[, matched$row_index, drop = FALSE]
  y <- matched$ct

  # Mean-impute missing genotypes per SNP.
  row_sum <- rowSums(G, na.rm = TRUE)
  row_obs <- rowSums(!is.na(G))
  row_mu <- row_sum / pmax(row_obs, 1)
  na_pos <- which(is.na(G), arr.ind = TRUE)
  if (nrow(na_pos) > 0) {
    G[na_pos] <- row_mu[na_pos[, 1]]
  }

  # Fast single-SNP linear test y ~ SNP (no additional covariates).
  n <- length(y)
  y_mean <- mean(y)
  y_ss <- sum((y - y_mean)^2)
  sx <- rowSums(G)
  sx2 <- rowSums(G * G)
  sxy <- as.vector(G %*% y)
  sxx <- sx2 - (sx^2 / n)
  sxy_center <- sxy - (sx * y_mean)

  r <- sxy_center / sqrt(sxx * y_ss)
  r[!is.finite(r)] <- 0
  t_stat <- r * sqrt((n - 2) / pmax(1e-12, 1 - r^2))
  p_val <- 2 * pt(-abs(t_stat), df = n - 2)
  beta <- sxy_center / sxx

  gwas_results <- tibble(
    chr = bim$chr,
    snp = bim$snp,
    bp = bim$bp,
    beta = beta,
    p = p_val
  ) %>%
    arrange(p)

  dir.create("output", showWarnings = FALSE)
  readr::write_csv(gwas_results, file.path("output", "cms_qpcr_gwas_results.csv"))

  cat("GWAS finished. Results saved to output/cms_qpcr_gwas_results.csv\n")
  head(gwas_results, 15)
}
```

```{r gwas_followup}
if (exists("gwas_results")) {
  best_chr12 <- gwas_results %>% filter(chr == "12") %>% slice_min(order_by = p, n = 1)
  best_chr27 <- gwas_results %>% filter(chr == "27") %>% slice_min(order_by = p, n = 1)

  best_chr12
  best_chr27

  # Estimate variance explained by the top SNP.
  top_snp <- gwas_results$snp[1]
  snp_idx <- match(top_snp, bim$snp)
  x <- plink_obj$X[snp_idx, matched$row_index]
  x[is.na(x)] <- mean(x, na.rm = TRUE)
  fit_top <- lm(y ~ x)
  tibble(
    top_snp = top_snp,
    beta = unname(coef(fit_top)[2]),
    r_squared = summary(fit_top)$r.squared
  )
}
```

# Simulation of Selection Response

```{r simulation_helpers}
build_genetic_cov <- function(h2, vp, rho_qh, rho_qs, rho_hs) {
  va <- h2 * vp
  R <- matrix(
    c(
      1, rho_qh, rho_qs,
      rho_qh, 1, rho_hs,
      rho_qs, rho_hs, 1
    ),
    nrow = 3, byrow = TRUE
  )

  G <- diag(sqrt(va)) %*% R %*% diag(sqrt(va))
  as.matrix(Matrix::nearPD(G)$mat)
}

simulate_selection <- function(
  index_name = c("I1_qPCR", "I2_histopathology", "I3_index"),
  n_rep = 100,
  n_generations = 10,
  n_families = 200,
  offspring_per_family = 20,
  n_candidates = 700,
  means = c(qPCR_CT = 27, Histopathology = 1.95, Survival = 0.90),
  h2 = c(qPCR_CT = 0.45, Histopathology = 0.45, Survival = 0.20),
  vp = c(qPCR_CT = 10, Histopathology = 0.8, Survival = 1),
  accuracies = c(qPCR_CT = 0.7, Histopathology = 0.8, Survival = 0.5),
  rho_qh = -0.8,
  rho_qs = 0.5,
  rho_hs = -0.6,
  seed = 1
) {
  index_name <- match.arg(index_name)
  set.seed(seed)

  G <- build_genetic_cov(h2, vp, rho_qh, rho_qs, rho_hs)
  va <- diag(G)
  err_sd <- sqrt(va * (1 / accuracies^2 - 1))

  n_sires <- n_families / 2
  n_dams <- n_families
  n_offspring <- n_families * offspring_per_family
  out <- vector("list", n_rep)

  for (r in seq_len(n_rep)) {
    # Initial candidate pool.
    tbv <- MASS::mvrnorm(n = n_candidates, mu = rep(0, 3), Sigma = G)
    rec <- matrix(NA_real_, nrow = n_generations + 1, ncol = 3)
    rec[1, ] <- colMeans(tbv)

    for (g in seq_len(n_generations)) {
      n <- nrow(tbv)
      sex <- sample(c("M", "F"), size = n, replace = TRUE)

      # Build EBV from TBV and target accuracy.
      ebv <- tbv + matrix(rnorm(n * 3), nrow = n, ncol = 3) %*% diag(err_sd)

      # Define selection score for each index.
      score <- switch(
        index_name,
        I1_qPCR = ebv[, 1],                                  # high CT is better
        I2_histopathology = -ebv[, 2],                       # low histopathology is better
        I3_index = as.numeric(scale(ebv[, 1]) - scale(ebv[, 2])) # combined index
      )

      male_idx <- which(sex == "M")
      female_idx <- which(sex == "F")
      if (length(male_idx) < n_sires || length(female_idx) < n_dams) {
        sex <- sample(rep(c("M", "F"), length.out = n))
        male_idx <- which(sex == "M")
        female_idx <- which(sex == "F")
      }

      sires <- male_idx[order(score[male_idx], decreasing = TRUE)][seq_len(n_sires)]
      dams <- female_idx[order(score[female_idx], decreasing = TRUE)][seq_len(n_dams)]

      # Nested mating: 1 sire mates with 2 dams.
      offspring_tbv <- matrix(NA_real_, nrow = n_offspring, ncol = 3)
      row_ptr <- 1L
      for (s in seq_len(n_sires)) {
        dams_this_sire <- dams[((2 * s) - 1):(2 * s)]
        for (d in dams_this_sire) {
          mid_parent <- matrix(
            (tbv[sires[s], ] + tbv[d, ]) / 2,
            nrow = offspring_per_family,
            ncol = 3,
            byrow = TRUE
          )
          mendelian <- MASS::mvrnorm(
            n = offspring_per_family,
            mu = rep(0, 3),
            Sigma = 0.5 * G
          )
          offspring_tbv[row_ptr:(row_ptr + offspring_per_family - 1), ] <- mid_parent + mendelian
          row_ptr <- row_ptr + offspring_per_family
        }
      }

      rec[g + 1, ] <- colMeans(offspring_tbv)

      # Keep a limited candidate set for the next generation.
      keep <- sample.int(n_offspring, size = n_candidates, replace = FALSE)
      tbv <- offspring_tbv[keep, , drop = FALSE]
    }

    out[[r]] <- tibble(
      generation = 0:n_generations,
      replicate = r,
      index = index_name,
      qPCR_CT = means["qPCR_CT"] + rec[, 1],
      Histopathology = pmax(0, means["Histopathology"] + rec[, 2]),
      Survival = pmin(1, pmax(0, means["Survival"] + rec[, 3]))
    )
  }

  bind_rows(out) %>%
    pivot_longer(cols = c(qPCR_CT, Histopathology, Survival), names_to = "trait", values_to = "value")
}
```

```{r simulation_run}
indices <- c("I1_qPCR", "I2_histopathology", "I3_index")

sim_results <- purrr::map_dfr(
  indices,
  ~ simulate_selection(
    index_name = .x,
    n_rep = params$n_replicates,
    n_candidates = params$n_candidates,
    seed = 100 + match(.x, indices)
  )
)

sim_summary <- sim_results %>%
  group_by(index, generation, trait) %>%
  summarise(
    mean = mean(value),
    p10 = quantile(value, 0.10),
    p90 = quantile(value, 0.90),
    .groups = "drop"
  )

sim_summary %>%
  filter(generation %in% c(0, 3, 6, 10)) %>%
  arrange(trait, index, generation)
```

```{r simulation_plot}
sim_summary %>%
  ggplot(aes(x = generation, y = mean, color = index, fill = index)) +
  geom_ribbon(aes(ymin = p10, ymax = p90), alpha = 0.14, color = NA) +
  geom_line(linewidth = 1) +
  facet_wrap(~trait, scales = "free_y", ncol = 1) +
  labs(
    title = "Simulated response to selection under three indices",
    x = "Generation",
    y = "Trait mean",
    color = "Index",
    fill = "Index"
  ) +
  theme_bw()
```

# Sensitivity Analysis

```{r sensitivity}
sensitivity_reps <- if (run_heavy_steps) params$n_replicates else min(30, params$n_replicates)

scenarios <- tribble(
  ~scenario, ~rho_qh, ~rho_qs, ~rho_hs,
  "Base", -0.8,  0.5, -0.6,
  "Moderate_qh", -0.6, 0.5, -0.6,
  "Lower_qs", -0.8, 0.3, -0.6,
  "Lower_hs", -0.8, 0.5, -0.4
)

sensitivity_results <- purrr::pmap_dfr(
  scenarios,
  function(scenario, rho_qh, rho_qs, rho_hs) {
    purrr::map_dfr(
      c("I1_qPCR", "I2_histopathology", "I3_index"),
      ~ simulate_selection(
        index_name = .x,
        n_rep = sensitivity_reps,
        n_candidates = params$n_candidates,
        rho_qh = rho_qh,
        rho_qs = rho_qs,
        rho_hs = rho_hs,
        seed = 1000 + match(.x, c("I1_qPCR", "I2_histopathology", "I3_index"))
      ) %>%
        mutate(scenario = scenario)
    )
  }
)

sensitivity_g3 <- sensitivity_results %>%
  filter(generation == 3) %>%
  group_by(scenario, index, trait) %>%
  summarise(mean_g3 = mean(value), .groups = "drop")

sensitivity_g3
```

# Interpretation Notes

-   This notebook reproduces the **workflow logic** from the PDF with available local data.
-   The qPCR GWAS usually shows a strong chr27 signal and a secondary chr12 signal in this dataset.
-   Simulation outputs are sensitive to assumptions not fully specified in the PDF (especially candidate pool size and index construction), so treat numeric agreement as approximate.
-   For closer replication, replace the simplified GWAS and simulation assumptions with the original internal script/model settings if available.
