---
title: "EDA"
author: "Junxiu"
date: "2026-02-28"
output: html_document
---

# Packages & Source functions

```{r library,warning=FALSE}
library(tidyverse)
library(readxl)
library(janitor)
library(skimr)
library(naniar)
library(stringr)
library(genio)
library(here)

source("function/load_functions.R")
source(here("script", "load_script.R"))

has_asreml <- requireNamespace("asreml", quietly = TRUE)
if (has_asreml) {
  library(asreml)
} else {
  message("Package 'asreml' is not available in this R session. ASReml chunks will be skipped.")
}
```

# File path

```{r path}
# phenotype
pheno_file1 <- file.path("data", "phenotype",
                         "CMS challenge Scotland 22G BE20041B sample IDs_with PCR11102022.xlsx")
pheno_file2 <- file.path("data", "phenotype",
                         "Data - CMS MOWI genetics project Scotland.xlsx")

# pedigree
ped_file <- file.path("data", "pedigree", "pedigree22G_CMSSalmoStrong.txt")

# genotype
bed_file <- file.path("data", "genotype", "2022CMSGeno_SalmoStrong.bed")
bim_file <- file.path("data", "genotype", "2022CMSGeno_SalmoStrong.bim")
fam_file <- file.path("data", "genotype", "2022CMSGeno_SalmoStrong.fam")
```

# EDA

## Phenotype

### Glimpse of the data

```{r}
pcr_pheno <- read_excel(pheno_file1, sheet = "Sheet2") %>%
  clean_names() %>%
  rename(
    fish_number = fish_nunber,
    test_replicate = test_replicate,
    tissue_type = tissue_type,
    reported_result = reported_result,
    cp_1 = cp_value_8,
    cp_2 = cp_value_9,
    cp_3 = cp_value_10
  ) %>%
  mutate(
    sample_id = as.factor(sample_id),
    reported_result = as.factor(reported_result),
    tissue_type = as.factor(tissue_type),
    species = as.factor(species)
  )
```

```{r}
head(pcr_pheno, 10)
```

```{r}
histo_pheno <- read_excel(pheno_file2, sheet = "Scotland samples", skip = 1) %>%
  clean_names() %>%
  rename(
    id = id,
    atrium_area_background = area_background_2,
    atrium_area_inflamed = area_inflamed_tissue,
    atrium_area_rbc = area_red_blood_cells_4,
    atrium_area_noninflamed = area_non_inflamed_tissue_5,
    atrium_area_investigated = area_investigated_6,
    atrium_pct_inflam = percentage_inflammation,
    atrium_rel_inflam = relative_area_inflammation_8,
    spong_area_background = area_background_11,
    spong_area_inflamed = area_inflammed_tissue,
    spong_area_rbc = area_red_blood_cells_13,
    spong_area_noninflamed = area_non_inflamed_tissue_14,
    spong_area_investigated = area_investigated_15,
    spong_pct_inflam = precentage_inflammation,
    spong_rel_inflam = relative_area_inflammation_17,
    ratio_inflam = ratio_inflammation_atrium_spongiosum,
    ratio_control = ratio_controll
  )
```

```{r}
head(histo_pheno, 10)
```

```{r}
glimpse(pcr_pheno)
```

```{r}
glimpse(histo_pheno)
```

### Summary of the data

```{r}
skim(pcr_pheno)
```

```{r}
skim(histo_pheno)
```

```{r}
# Whether there are replicate of ID and fish number
pcr_pheno %>% count(sample_id) %>% filter(n > 1)
pcr_pheno %>% count(fish_number) %>% filter(n > 1)
histo_pheno %>% count(id) %>% filter(n > 1)
# No!
```

### Distribution of Phenotype

```{r}
pcr_pheno %>%
  pivot_longer(cols = c(cp_1, cp_2, cp_3), names_to = "rep", values_to = "cp") %>%
  ggplot(aes(x = cp)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~rep, scales = "free_y") +
  theme_bw()

```

Take a look at the `mean` and `median` of `Cp_value`

```{r}
pcr_pheno %>%
  mutate(
    cp_mean   = rowMeans(across(c(cp_1, cp_2, cp_3)), na.rm = TRUE),
    cp_median = apply(across(c(cp_1, cp_2, cp_3)), 1, median, na.rm = TRUE)
  ) %>%
  select(cp_mean, cp_median) %>%
  pivot_longer(cols = everything(), names_to = "summary_type", values_to = "cp") %>%
  ggplot(aes(x = cp)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 fill = "steelblue",
                 color = "white") +
  geom_density(color = "red", linewidth = 1) +
  facet_wrap(~summary_type, scales = "free") +
  theme_bw()
```

We could see a little left skew for both mean and median value of the `Cp_value`.

```{r}
pcr_pheno %>%
  mutate(
    cp_mean   = rowMeans(across(c(cp_1, cp_2, cp_3)), na.rm = TRUE),
    cp_median = apply(across(c(cp_1, cp_2, cp_3)), 1, median, na.rm = TRUE)
  ) %>%
  select(cp_mean, cp_median) %>%
  pivot_longer(everything(), names_to = "trait", values_to = "cp") %>%
  ggplot(aes(sample = cp)) +
  stat_qq(size = 2, alpha = 0.6) +
  stat_qq_line(color = "red", linewidth = 1) +
  facet_wrap(~trait) +
  theme_bw() +
  labs(
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  )
```

The normal distribution assumption seems acceptable.

### Distribution of histopathology

```{r}
histo_pheno %>%
  select(atrium_pct_inflam, spong_pct_inflam, ratio_inflam) %>%
  pivot_longer(everything(), names_to = "trait", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 fill = "darkgrey",
                 color = "white") +
  geom_density(color = "red", linewidth = 1) +
  facet_wrap(~trait, scales = "free") +
  theme_bw()
```

Consider the data transformation: take `log` value.

```{r warning=FALSE}
histo_pheno %>%
  select(atrium_pct_inflam, spong_pct_inflam, ratio_inflam) %>%
  pivot_longer(everything(), names_to = "trait", values_to = "value") %>%
  mutate(value = log(value)) %>%
  ggplot(aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 fill = "darkgrey",
                 color = "white") +
  geom_density(color = "red", linewidth = 1) +
  facet_wrap(~trait, scales = "free") +
  theme_bw()
```

```{r}
# Calculate cp mean, sd
# Calculate inflam differ and mean

pcr_pheno <- pcr_pheno %>%
  mutate(
    cp_mean = rowMeans(select(., cp_1, cp_2, cp_3), na.rm = TRUE),
    cp_sd = apply(select(., cp_1, cp_2, cp_3), 1, sd, na.rm = TRUE)
  )

histo_pheno <- histo_pheno %>%
  mutate(
    inflam_diff = atrium_pct_inflam - spong_pct_inflam,
    inflam_mean = (atrium_pct_inflam + spong_pct_inflam) / 2
  )

summary(pcr_pheno$cp_mean)
cat("\n")
summary(pcr_pheno$cp_sd)
cat("\n")
summary(histo_pheno$inflam_diff)
```

From the result above, the max of $\text{cp_sd}$ value could be 21 which is a little bit large which can be think of as outlier. Double check:

```{r}
pcr_pheno %>%
    arrange(desc(cp_sd)) %>% 
    select(sample_id, sample_no, cp_1, cp_2, cp_3, cp_mean, cp_sd) %>% 
    head(20)
    
```

We can see that $\text{sample_ID}$ which 420500 and 418601 have abnormal $\text{cp_value}$. According to Panya's suggestion, we just treat as 0 and take the mean in the nest step.

Now, we look at the $\text{histo_pheno}$ difference between two areas:

```{r}
histo_pheno %>%
  ggplot(aes(x = inflam_diff)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  theme_bw()

histo_pheno %>%
  arrange(desc(inflam_diff)) %>%
  select(id, atrium_pct_inflam, spong_pct_inflam, inflam_diff) %>%
  head(50)

```

We notice that there are 6 NAs, now we find them:

```{r}
histo_pheno %>%
  summarise(
    n_missing_atrium = sum(is.na(atrium_pct_inflam)),
    n_missing_spong = sum(is.na(spong_pct_inflam)),
    n_missing_diff = sum(is.na(inflam_diff))
  )

histo_pheno %>%
  filter(is.na(inflam_diff)) %>%
  select(id, atrium_pct_inflam, spong_pct_inflam)
```

We find that 2022_0221_115, 130, 375, 383, 431, 491 are missing $\text{atrium_pct_inflam}$ value.

### Join the phenotype IDs

```{r}
pcr_pheno2 <- pcr_pheno %>%
  mutate(
    batch = str_extract(sample_no, "\\d{4}-\\d{4}"),            
    fish_n = as.integer(str_extract(sample_no, "(?<=-F)\\d+")), 
    fish3  = str_pad(fish_n, width = 3, pad = "0"),            
    id_key = paste0(str_replace_all(batch, "-", "_"), "_", fish3)  
  )


histo_pheno2 <- histo_pheno %>%
  mutate(
    batch = str_extract(id, "\\d{4}_\\d{4}"),              
    fish_n = as.integer(str_extract(id, "(?<=_)\\d+(?=\\s*\\()")), 
    fish3 = str_pad(fish_n, 3, pad = "0"),
    id_key = paste0(batch, "_", fish3)
  )

pheno_joined <- pcr_pheno2 %>%
  left_join(histo_pheno2, by = "id_key")
```

We now get the **joined form** $\text{pheno_joined}$

```{r}
head(pheno_joined, 20)
```

### Build phenotype-to-pedigree (from Sheet1)

`Sheet1` contains `animal/sire/dam` fields and is used as the bridge from phenotype IDs to pedigree IDs.

```{r}
normalize_cms_id <- function(x) {
  x <- as.character(x)
  x[x %in% c("", "NA", "NaN", "0", "2")] <- NA_character_

  # Animal IDs in Sheet1 miss "00" after ICM
  x <- str_replace(x, "^2022IRS1ICMFR", "2022IRS1ICM00FR")

  # Parent IDs in Sheet1 miss "0000" after CBN to fit pedigree
  x <- str_replace(x, "^2018IRS1CBN([0-9]{5}[A-Z][0-9]{2})$", "2018IRS1CBN0000\\1")
  x
}

sheet1_map <- read_excel(pheno_file1, sheet = "Sheet1") %>%
  clean_names() %>%
  transmute(
    sample_id = as.character(sample_id),
    fish_no = suppressWarnings(as.integer(fish_no)),
    histo_cassette_id = histo_cassette_id,
    histo_key = if_else(
      !is.na(histo_cassette_id),
      paste0(
        str_replace_all(str_extract(histo_cassette_id, "^\\d{4}/\\d{4}"), "/", "_"),
        "_",
        str_pad(as.integer(str_extract(histo_cassette_id, "\\d+$")), 3, pad = "0")
      ),
      NA_character_
    ),
    genotyping_tube_id = genotyping_tube_id,
    animal = normalize_cms_id(animal),
    sire = normalize_cms_id(sire),
    dam = normalize_cms_id(dam)
  ) %>%
  distinct(sample_id, .keep_all = TRUE)

pheno_ped <- pheno_joined %>%
  mutate(sample_id = as.character(sample_id)) %>%
  left_join(sheet1_map, by = "sample_id")
```

```{r}
head(pheno_ped, 50)
```

## Pedigree Information

```{r}
pedigree <- read_table(ped_file, col_types = cols(
  id = col_character(),
  sire = col_character(),
  dam = col_character(),
  sibgroup = col_character(),
  YC = col_double()
)) %>%
  mutate(
    sire = na_if(sire, "0"),
    dam = na_if(dam, "0"),
    is_founder = is.na(sire) & is.na(dam),
    n_known_parents = (!is.na(sire)) + (!is.na(dam))
  )
```

Have a look at the pedigree structure:

```{r}
head(pedigree, 20)
tail(pedigree, 20)
```

Some summary of pedigree information:

```{r}
qgm_ped_summary(pedigree[,1:3])
```

From the result, we could see that there are 3051 different individuals, no multiple IDs. Quite clean!

Quite a lot missing parents information.

```{r}
# Some summary statistics about pedigree
pedigree %>%
  summarise(
    n_animals = n(),
    n_unique_id = n_distinct(id),
    n_founders = sum(is_founder),
    n_non_founders = sum(!is_founder),
    n_two_known_parents = sum(n_known_parents == 2),
    n_one_known_parent = sum(n_known_parents == 1),
    n_no_known_parent = sum(n_known_parents == 0)
  )

pedigree %>%
  count(sibgroup, sort = TRUE) %>%
  head(15)

pedigree %>%
  count(YC, sort = TRUE)
```

## Fitting Sire Model

Deal with A-inverse matrix:

```{r}
ainv <- build_ainv_asreml(ped_file)
ainv_ids <- attr(ainv, "rowNames")
n_unique(ainv_ids)
```

```{r}
head(pheno_ped, 500)
nrow(pheno_ped)

tail(pheno_ped, 100)
```

```{r eval=FALSE}
# Prepare the dataset
sire_model_dat <- pheno_ped %>%
  mutate(
    sire = as.character(sire),
    cp_1_clean = na_if(cp_1, 0),
    cp_2_clean = na_if(cp_2, 0),
    cp_3_clean = na_if(cp_3, 0),
    cp_mean = rowMeans(cbind(cp_1_clean, cp_2_clean, cp_3_clean), na.rm = TRUE),
    cp_mean = if_else(is.nan(cp_mean), NA_real_, cp_mean)
  ) %>%
  filter(
    !is.na(cp_mean),
    !is.na(sire),
    sire %in% ainv_ids
  ) %>%
  mutate(
    sire = factor(sire, levels = ainv_ids)
  )

```

If you run `dim(sire_model_dat)`, there are only 469 rows which because `sire %in% ainc_ids` delete some **unknown** sires in `ainv` matrix.

```{r}
dim(sire_model_dat)
tail(sire_model_dat, 20)
```

Model:

```{r}
fit_sire <- asreml(
  fixed = cp_mean ~ 1,
  random = ~ vm(sire, ainv),
  residual = ~ idv(units),
  data = sire_model_dat
)
```

```{r}
qgm_model_summary(fit_sire)
```

From the result, we can see the model converged and `log-likelihood` stabilized at `-724` .

#### **Variance Components:** 

`vm(sire,ainv)` is the estimate of sire variance components and residual is estimated by `units!units`.

#### **Fixed effect:** 

Only Intercept effect.

#### Predicted Transmitting Abilities/PTAs:

The estimates obtained from fitting a Sire Model are the predicted superiority or inferiority of the offspring of the sires, these are called **Predicted Transmitting Abilities** or PTAs.

```{r warning=FALSE}
pta <- qgm_random_effects_pta(fit_sire, "sire")
head(pta, 30)
tail(pta, 30)
```

#### Merge PTAs with whole pedigree information

```{r}
colnames(pta)[1] <- "pta"
colnames(pta)[2] <- "se"
colnames(pta)[4] <- "id" # rename pta$levels as pta$id to match with ped data frame ready for merging
pedx <- merge(pedigree, pta, by = "id")
head(pedx, 200)
```

```{r}
unique(pedx$sibgroup)
# PTA 
hist(pedx[pedx$sibgroup %in% "S1CBN",]$pta,breaks=20)
hist(pedx[pedx$sibgroup %in% "S1CBI",]$pta,breaks=20)
hist(pedx[pedx$sibgroup %in% "S1ICM",]$pta,breaks=20)

```

Option model:

```{r eval=FALSE}
# Optional model: use histopathology as y and Cp as fixed covariate
histo_model_dat <- sire_base %>%
  left_join(
    histo_pheno2 %>% select(id_key, atrium_pct_inflam, spong_pct_inflam, ratio_inflam),
    by = c("histo_key" = "id_key")
  ) %>%
  filter(
    !is.na(atrium_pct_inflam),
    !is.na(cp_mean_clean),
    !is.na(sire_final),
    sire_final %in% ainv_ids
  ) %>%
  mutate(
    sire = factor(sire_final, levels = ainv_ids)
  )

fit_inflam <- asreml(
  fixed = atrium_pct_inflam ~ cp_mean_clean,
  random = ~ vm(sire, Ainv),
  residual = ~ idv(units),
  data = histo_model_dat
)
```

## Read Genotype file

```{r}
# Read Plink files (bim/fam/bed)
plink_obj <- read_plink(tools::file_path_sans_ext(bed_file))
geno <- plink_obj$X
bim <- plink_obj$bim
fam <- plink_obj$fam
```

```{r}
head(fam, 10)
head(bim, 10)
tail(fam, 10)
tail(bim, 10)
```

```{r}
dim(geno)
dim(fam)
dim(bim)
```

```{r}
sum(fam$id %in% pedigree$id)
sum(pedigree$id %in% fam$id)
```

All genotyped samples can be matched to pedigree IDs.

```{r}
fam_ped <- fam %>%
  left_join(
    pedigree %>% select(id, sire, dam, sibgroup, YC, is_founder, n_known_parents),
    by = "id"
  )

fam_ped %>%
  summarise(
    n_fam = n(),
    n_missing_in_pedigree = sum(is.na(sibgroup)),
    n_founders_in_genotyped = sum(is_founder, na.rm = TRUE),
    n_non_founders_in_genotyped = sum(!is_founder, na.rm = TRUE)
  )

fam_ped %>%
  count(sibgroup, sort = TRUE) %>%
  head(15)
```

## Genotype QC Summary

```{r}
snp_call_rate <- 1 - rowMeans(is.na(geno))
sample_call_rate <- 1 - colMeans(is.na(geno))

alt_af <- rowMeans(geno, na.rm = TRUE) / 2
maf <- pmin(alt_af, 1 - alt_af)

sample_qc <- tibble(
  id = fam$id,
  call_rate = sample_call_rate,
  het_rate = colMeans(geno == 1, na.rm = TRUE),
  hom_ref_rate = colMeans(geno == 0, na.rm = TRUE),
  hom_alt_rate = colMeans(geno == 2, na.rm = TRUE)
)

snp_qc <- tibble(
  snp_id = bim$id,
  chr = bim$chr,
  pos = bim$pos,
  call_rate = snp_call_rate,
  alt_af = alt_af,
  maf = maf
)
```

```{r}
sample_qc %>%
  summarise(
    n_samples = n(),
    mean_call_rate = mean(call_rate),
    min_call_rate = min(call_rate),
    mean_het_rate = mean(het_rate),
    sd_het_rate = sd(het_rate)
  )

snp_qc %>%
  summarise(
    n_snps = n(),
    mean_call_rate = mean(call_rate),
    min_call_rate = min(call_rate),
    n_monomorphic = sum(maf == 0),
    n_low_maf_001 = sum(maf < 0.01),
    n_low_maf_005 = sum(maf < 0.05)
  )
```

```{r}
sample_qc %>%
  ggplot(aes(x = het_rate)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  theme_bw() +
  xlab("Sample heterozygosity (proportion of genotype == 1)")

snp_qc %>%
  ggplot(aes(x = maf)) +
  geom_histogram(bins = 50, fill = "darkolivegreen3", color = "white") +
  theme_bw() +
  xlab("Minor allele frequency")

snp_qc %>%
  ggplot(aes(x = call_rate)) +
  geom_histogram(bins = 30, fill = "tomato", color = "white") +
  theme_bw() +
  xlab("SNP call rate")
```

## Phenotype Coverage Check

```{r}
pcr_ids <- pcr_pheno %>%
  mutate(fish_n = as.integer(str_extract(sample_no, "(?<=-F)\\d+"))) %>%
  distinct(fish_n) %>%
  mutate(in_pcr = TRUE)

histo_ids <- histo_pheno %>%
  mutate(fish_n = as.integer(str_extract(id, "(?<=_)\\d+(?=\\s*\\()"))) %>%
  distinct(fish_n) %>%
  mutate(in_histo = TRUE)

pheno_overlap <- full_join(pcr_ids, histo_ids, by = "fish_n") %>%
  mutate(
    in_pcr = replace_na(in_pcr, FALSE),
    in_histo = replace_na(in_histo, FALSE),
    overlap_group = case_when(
      in_pcr & in_histo ~ "Both PCR and Histo",
      in_pcr & !in_histo ~ "PCR only",
      !in_pcr & in_histo ~ "Histo only"
    )
  )

pheno_overlap %>%
  count(overlap_group)
```

```{r}
pheno_overlap %>%
  ggplot(aes(x = overlap_group, fill = overlap_group)) +
  geom_bar(show.legend = FALSE) +
  theme_bw() +
  ylab("Number of fish IDs") +
  xlab(NULL)
```

```{r}
sum(as.character(pcr_pheno$sample_id) %in% fam$id)
sum(pcr_pheno$sample_no %in% fam$id)
sum(histo_pheno$id %in% fam$id)
```

No direct string overlap is observed between phenotype IDs and PLINK `fam$id`. In this project, `Sheet1` serves as the external crosswalk to connect phenotype and pedigree/genotype IDs.
