---
title: "EDA"
author: "Junxiu"
date: "2026-02-28"
output: html_document
---

# Packages

```{r library,warning=FALSE}
library(tidyverse)
library(readxl)
library(janitor)
library(skimr)
library(naniar)
library(data.table)
library(stringr)

```

# File path

```{r path}
# phenotype
pheno_file1 <- file.path("data", "phenotype",
                         "CMS challenge Scotland 22G BE20041B sample IDs_with PCR11102022.xlsx")
pheno_file2 <- file.path("data", "phenotype",
                         "Data - CMS MOWI genetics project Scotland.xlsx")

# pedigree
ped_file <- file.path("data", "pedigree", "pedigree22G_CMSSalmoStrong.txt")

# genotype
bed_file <- file.path("data", "genotype", "2022CMSGeno_SalmoStrong.bed")
bim_file <- file.path("data", "genotype", "2022CMSGeno_SalmoStrong.bim")
fam_file <- file.path("data", "genotype", "2022CMSGeno_SalmoStrong.fam")
```

# EDA

## Phenotype

### Glimpse of the data

```{r}
pcr_pheno <- read_excel(pheno_file1, sheet = "Sheet2") %>%
  clean_names() %>%
  rename(
    fish_number = fish_nunber,
    test_replicate = test_replicate,
    tissue_type = tissue_type,
    reported_result = reported_result,
    cp_1 = cp_value_8,
    cp_2 = cp_value_9,
    cp_3 = cp_value_10
  ) %>%
  mutate(
    reported_result = as.factor(reported_result),
    tissue_type = as.factor(tissue_type),
    species = as.factor(species)
  )
```

```{r}
head(pcr_pheno, 10)
```

```{r}
histo_pheno <- read_excel(pheno_file2, sheet = "Scotland samples", skip = 1) %>%
  clean_names() %>%
  rename(
    id = id,
    atrium_area_background = area_background_2,
    atrium_area_inflamed = area_inflamed_tissue,
    atrium_area_rbc = area_red_blood_cells_4,
    atrium_area_noninflamed = area_non_inflamed_tissue_5,
    atrium_area_investigated = area_investigated_6,
    atrium_pct_inflam = percentage_inflammation,
    atrium_rel_inflam = relative_area_inflammation_8,
    spong_area_background = area_background_11,
    spong_area_inflamed = area_inflammed_tissue,
    spong_area_rbc = area_red_blood_cells_13,
    spong_area_noninflamed = area_non_inflamed_tissue_14,
    spong_area_investigated = area_investigated_15,
    spong_pct_inflam = precentage_inflammation,
    spong_rel_inflam = relative_area_inflammation_17,
    ratio_inflam = ratio_inflammation_atrium_spongiosum,
    ratio_control = ratio_controll
  )
```

```{r}
head(histo_pheno, 10)
```

```{r}
glimpse(pcr_pheno)
```

```{r}
glimpse(histo_pheno)
```

### Summary of the data

```{r}
skim(pcr_pheno)
```

```{r}
skim(histo_pheno)
```

```{r}
# Whether there are replicate of ID and fish number
pcr_pheno %>% count(sample_id) %>% filter(n > 1)
pcr_pheno %>% count(fish_number) %>% filter(n > 1)
histo_pheno %>% count(id) %>% filter(n > 1)
# No!
```

### Distribution of Phenotype

```{r}
pcr_pheno %>%
  pivot_longer(cols = c(cp_1, cp_2, cp_3), names_to = "rep", values_to = "cp") %>%
  ggplot(aes(x = cp)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~rep, scales = "free_y") +
  theme_bw()

pcr_pheno %>%
  ggplot(aes(x = reported_result)) +
  geom_bar(fill = "tomato") +
  theme_bw()

pcr_pheno %>%
  ggplot(aes(x = tissue_type, fill = reported_result)) +
  geom_bar(position = "fill") +
  theme_bw() +
  ylab("Proportion")
```

### Distribution of histopathology

```{r}
histo_pheno %>%
  select(atrium_pct_inflam, spong_pct_inflam, ratio_inflam) %>%
  pivot_longer(everything(), names_to = "trait", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "darkseagreen3", color = "white") +
  facet_wrap(~trait, scales = "free") +
  theme_bw()
```

```{r}
# Calculate cp mean, sd
# Calculate inflam differ and mean

pcr_pheno <- pcr_pheno %>%
  mutate(
    cp_mean = rowMeans(select(., cp_1, cp_2, cp_3), na.rm = TRUE),
    cp_sd = apply(select(., cp_1, cp_2, cp_3), 1, sd, na.rm = TRUE)
  )

histo_pheno <- histo_pheno %>%
  mutate(
    inflam_diff = atrium_pct_inflam - spong_pct_inflam,
    inflam_mean = (atrium_pct_inflam + spong_pct_inflam) / 2
  )

summary(pcr_pheno$cp_mean)
cat("\n")
summary(pcr_pheno$cp_sd)
cat("\n")
summary(histo_pheno$inflam_diff)
```

From the result above, the max of $\text{cp_sd}$ value could be 21 which is a little bit large which can be think of as outlier. Double check:

```{r}
pcr_pheno %>%
    arrange(desc(cp_sd)) %>% 
    select(sample_id, sample_no, cp_1, cp_2, cp_3, cp_mean, cp_sd) %>% 
    head(20)
    
```

We can see that $\text{sample_ID}$ which 420500 and 418601 have abnormal $\text{cp_value}$.

Two proposed methods:

1.  change 0 to 'NA'
2.  See reported_result as categorical variable?

Now, we look at the $\text{histo_pheno}$:

```{r}
histo_pheno %>%
  ggplot(aes(x = inflam_diff)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  theme_bw()

histo_pheno %>%
  arrange(desc(inflam_diff)) %>%
  select(id, atrium_pct_inflam, spong_pct_inflam, inflam_diff) %>%
  head(50)

```

We notice that there are 6 NAs, now we find them:

```{r}
histo_pheno %>%
  summarise(
    n_missing_atrium = sum(is.na(atrium_pct_inflam)),
    n_missing_spong = sum(is.na(spong_pct_inflam)),
    n_missing_diff = sum(is.na(inflam_diff))
  )

histo_pheno %>%
  filter(is.na(inflam_diff)) %>%
  select(id, atrium_pct_inflam, spong_pct_inflam)
```

We find that 2022_0221_115, 130, 375, 383, 431, 491 are missing $\text{atrium_pct_inflam}$ value.

### Join the phenotype IDs

```{r}
pcr_pheno2 <- pcr_pheno %>%
  mutate(
    batch = str_extract(sample_no, "\\d{4}-\\d{4}"),            
    fish_n = as.integer(str_extract(sample_no, "(?<=-F)\\d+")), 
    fish3  = str_pad(fish_n, width = 3, pad = "0"),            
    id_key = paste0(str_replace_all(batch, "-", "_"), "_", fish3)  
  )


histo_pheno2 <- histo_pheno %>%
  mutate(
    batch = str_extract(id, "\\d{4}_\\d{4}"),              
    fish_n = as.integer(str_extract(id, "(?<=_)\\d+(?=\\s*\\()")), 
    fish3 = str_pad(fish_n, 3, pad = "0"),
    id_key = paste0(batch, "_", fish3)
  )

pheno_joined <- pcr_pheno2 %>%
  left_join(histo_pheno2, by = "id_key")
```

We now get the **joined form** $\text{pheno_joined}$

```{r}
head(pheno_joined, 20)
```

## Pedigree Information

```{r}
pedigree <- read_table2(ped_file, col_types = cols(
    id = col_character(),
    sire = col_character(),
    dam = col_character(),
    sibgroup = col_character(),
    YC = col_double()
    ))
```

Have a look at the pedigree structure:

```{r}
head(pedigree, 20)
tail(pedigree, 20)
```
