---
title: "Tutorial 5: Aviagen Data"
author: "QGM Team"
date: '`r format(Sys.time(), "%H:%M:%S on %d %B, %Y")`'
output: 
  html_document:
    number_sections: TRUE
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Some libraries that have been met before. The QGM helper functions have not changed from Tutorial 4.

```{r}
library("asreml")
library("ggplot2")
library("GGally")
library("gridExtra")
source("qgm_helpers_t5.R")
```

# Background

The data today is real data from Aviagen plc, a global breeding company for chickens. The QGM Team thank Aviagen for providing this data to us. It concerns the body weight of chicks at 35 days and egg production of the mature female in a single line over 6 years. Some themes of the tutorial are:

-   All models are wrong but some models are more wrong than others
-   A genetic model is only as good as its environmental model

# Objective

The **objective** for the analysis is to estimate:

i.  the heritability of each trait;
ii. the additive genetic correlation between the two traits ($r_A$); and
iii. the environmental correlation between the two traits.

The **primary task for the workshop** is to decide what model to fit.

# Data and pedigree files

The **data** is in file `gen_avi.csv` and has the following information (missing is coded *`NA`*):

i.  *id* self-explanatory
ii. *sire* self-explanatory
iii. *dam* self-explanatory
iv. *sex* male 1, female 2
v.  *pou* this will **not** be used     
vi. *hw* hatch week of chick
vii. *dage* age of the individual's dam in weeks when it was hatched
viii. *bwt* body weight at 35 days in grams
ix. *eggs* total number of eggs laid.

```{r}
gen <- read.csv(file="gen_avi.csv",header=TRUE,sep=",",na.strings=c("NA"))
head(gen)
```

The **pedigree** is in the file `ped_avi.csv` and has the following information (missing is coded *`NA`*) :

i.  *id* coded as in data
ii. *sire* the *id* of the sire
iii. *dam* the *id* of the dam
iv. *use* appears as sire 1, as dam 2, and not a parent 0.

```{r}
ped <- read.csv(file="ped_avi.csv",header=TRUE,sep=",",na.strings=c("NA"))
head(ped, 100)
```

# Some starter code ...

```{r warning= FALSE, message= FALSE, progress= FALSE}
ggpairs(data = gen[,c("hw","dage","bwt","eggs")], diag= list(continuous= "bar",
        breaks='FD'), lower= list(continuous= wrap(ggally_points, size= 0.01,
        alpha= 0.1)))
```

**Q1.** What notable features are evident in the plots above?

Look at `hw` ...

```{r}
length(unique(gen$hw))
```

There are two columns that may be categorical `sex` and `dage` and the counts are easily tabulated.

```{r}
addmargins(table(gen$sex,gen$dage))
```

**Q2.** What notable features are evident in the table above?

**Q3.** How would you fit `dage` in the model?

The code below shows the number of non-missing records for each column of the data frame.

-   `is.na(df)` tests every element of data frame `df` for `NA`, with `NA` being TRUE
-   `!is.na()` negates the answer so `NA` is now FALSE
-   R will interpret TRUE as 1 and FALSE as 0, so `colSums(!is.na(gen))` counts the number of rows that are not missing for every column.

```{r}
colSums(!is.na(gen))
```

**Q4.** Anything to note about missing records?

```{r}
summary(gen$bwt)
sd(gen$bwt, na.rm=TRUE)
summary(gen$eggs)
sd(gen$eggs, na.rm=TRUE)
```

**Q5.** Is there anything suspicious about the number of eggs laid by the females? (Hint: *look at the minimum*)

```{r}
hist(gen$bwt,breaks='FD')
hist(gen$eggs,breaks='FD')
```

**Q6.** Anything new? (Maybe already noted in Q1).

```{r}
qgm_ped_summary(gen)
```

**Q7.** What notable information does this show?

**Q8.** What is your model be?

```{r}
ainv <- ainverse(ped[,1:3])
gen$id <- factor(gen$id)
gen$sex <- factor(gen$sex)
gen$jhw <- gen$hw
gen$jhw <- factor(gen$jhw)

```

```{r}
asr <- asreml(
  fixed = cbind(bwt, eggs) ~
          trait
        + at(trait,1):sex # only sex effect trait 1 (bwt)
        + trait:pol(dage,8)
        + trait:hw,

  random = ~ 
          corgh(trait):vm(id,ainv) # corgh: 2traits should have different variance and covariance
        + idh(trait):spl(hw) # spl: spline function. mdoel the non-linear hatch-week effect; idh: identity heterogeneous, cov() = 0.
        + corgh(trait):jhw, # the environemnt effect that represent same hatch-week effect

  residual = ~ id(units):corgh(trait),

  data = gen,
  maxit= 30, 
  workspace= "512mb"
)
```

**Q9.** What are your estimates?
